Towards planning of infantile cardiac surgery based on simulation

### 2012-02-27 Doc/Miccai/shell-model.tex
miccai 2012: shell model and method sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@339 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Several different models are commonly used for modelling thin structures or
tubular-like structures resembling blood vessels. \tomas{Add refs. or
reformulate!}
We choose to use thin shell elements for the predictive
simulation of the surgery. Shell elements are often considered numerically
too expensive, this is however not necessarily true.

The formulation of the thin shell elements is based on the mechanics of
continuum and these elements are commonly used in physics and mechanics for
deformation modelling of thin structures.

\tomas{TODO}
\noindent
briefly, refer to some textbook,
Eq. showing dependence on rest position:
simple $\myvec{f} = K(\myvec{x} - \myvec{x_0})$

\ldots

@@

Simulation using shell model satisfies all these requirements very well.

\tomas{TODO}
\noindent Test with tube + convergence
\begin{itemize}
  \item experimented with Olivier's model \cite{Comas2010b,Comas2010c} and
  our own model, in principle the shell implementation is not that
  important
  \item geometrical tool, independence on physical parameters -- E
\end{itemize}

>>>>

@@

Several different models are commonly used for modelling thin structures or
tubular-like structures resembling blood vessels. \tomas{Add refs. or
reformulate!} For the predictive simulation of the surgery we choose to use
thin shell elements, because in our opinion it satisfies all the
requirements very well. They are very often considered numerically too
expensive, this is however not necessarily true.

\tomas{image: single shell}

The formulation of the thin shell elements is based on the mechanics of
continuum. These elements are commonly used in physics and mechanics for
deformation modelling of thin structures because they are numerically more
stable then volumetric FEM methods in this situation. The shell element is
a 2-dimensional element whose thickness is a parameter of the simulation.
Often the thickness is considered constant over the surface of the element.
Shell combines forces from in-plane deformations, like stretching or
shearing, with bending forces. Such internal forces $\myvec{f}$ are a
function of current displacement measured as difference of current shape
$\myvec{x}$ from the rest shape $\myvec{x_0}$.

\begin{equation}
  \myvec{f} = K(\myvec{x} - \myvec{x_0})
\end{equation}

For more thorough description we refer the reader to the textbook about
Finite Elements for example \tomas{add ref}.

In the rest of the section we want to show that the performance of a model
based on the shell element is adequate for the simulation.

\begin{figure}[tbh]
  \begin{minipage}[b]{0.3\linewidth}
      \framebox[1cm]{\rule{0pt}{1cm}}
      %\includegraphics[width=\columnwidth]{img/}
      \caption{Bending shell element}
      \label{fig-shell}
  \end{minipage}
  \hspace{0.1\columnwidth}
  \begin{minipage}[b]{0.6\linewidth}
    \centering
    \begin{tabular}{cc}
     \includegraphics[width=0.5\columnwidth]{img/compare-bend.png}
      &
      \includegraphics[width=0.5\columnwidth]{img/compare-twist.png}
      \\
      \includegraphics[width=0.5\columnwidth]{img/compare-bend-other.png}
      &
      \includegraphics[width=0.5\columnwidth]{img/compare-twist-other.png}
    \end{tabular}
    \caption{Bending and twisting deformations. Upper: shell model; middle:
    artificial tube, bottom: Cosserat rod \cite{Li2009}. (Credits for the
    bottom images: Li et al.)}
    \label{fig-deformations}
  \end{minipage}
\end{figure}

Figure \ref{fig-deformations} shows different deformations that are likely to
occur during the surgery (bending and twisting). The figure compares our
results with the results of Li et al. \cite{Li2009}. As you can see the
results are closer to the real deformations.

The bending property of the shell allows us to use smaller number of
elements for the simulation and use high polygonal model only for
visualisation. The convergence of the model is a key factor here. In our experiments we
were able to use only as few as eight nodes around the circumference of a
tubular model while still maintaining realistic behaviour. Figure
\tomas{add-ref} shows the results for different number of nodes along the
circumference.

\tomas{image: convergence}

We have used the model of Olivier Comas \cite{Comas2010b,Comas2010c} and
also our own model. But our method described in section \ref{sec-method}
does not depend on any particular model. Since we use the method as a
geometrical tool we have also certain independence on physical parameters.
For example the variation of Young's modulus in the interval from $10^3$ to
$10^7$ resulted to only $0.1\%$ change of the deformation.

% vim: et sw=2 tw=75 fdm=marker fdc=2 spell

### 2012-02-27 Doc/Miccai/simulation.tex
add figure
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@338 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
miccai 2012: shell model and method sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@339 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\noindent surgical low-level tasks done with rest position trick (joining, patch, complex cutting/suturing)
\begin{itemize}
  \item describe the particular tasks and show that the essence is the same,
  \item describe our method with all the tricks
  \item include a schematic illustration
\end{itemize}

>>>>

There are several types of low-level task the surgeon usually employs while
performing the surgery.

One common procedure is reconnection and joining of blood vessels. Another
is fixing the stenosis of aorta where the three most common types of fixing
the problem are:

\begin{description}

  \item[End to end aortoplasty:] the surgeon completely cuts away the
    affected part of aorta and sutures the two ends of aorta back together.

  \item[Patch aortoplasty:] the surgeon performs a cut along the axis of
    the aorta thus allowing the narrow place to expand. Then applies a
    patch from artificial material to fill the hole in.

  \item[Subclavian flap aortoplasty:] The surgeon cuts the nearest artery,
    removing also half of the artery (along the axis) and part of the
    aorta in the affected place. He then uses the rest of the artery to
    patch the aorta.

\end{description}

From our point of view the all these procedures share the same similarity:
parts of the circulatory system are sutured together or to a patch. These
parts are not too far away from each other. On this we have based our
general approach in the simulation.

The simple way how to perform the suturing would be to connect the
respective edges with springs. This approach however adds some unwanted
energy to the system. Our method is based solely on the relaxation of the
mesh of shell elements.

We start with a mesh representing the situation after all the cuts have
been performed but before suturing takes place. In the first step we select
on the mesh edges that we want to suture together and join the
corresponding nodes together. The result is topologicaly different mesh
that has same number of elements. The Figure. \ref{fig-JoiningVessels} explains
this operation. Then we start the simulation on the modified mesh while
using the original mesh to define rest shapes of the shell elements. After
the mesh relaxes and the simulation reaches energetical equilibrium the
simulation terminates.

\begin{figure}[tbh]
\begin{center}
\includegraphics[width=\columnwidth]{img/rest_shape_scheme.png}
\end{center}

\caption{Scheme showing the method for joining two vessels.}
\label{fig-JoiningVessels}
\end{figure}

If the nodes that are to be sutured together are not in a close vicinity we
may experience unwanted quick increase of internal forces at the beginning
of the simulation. To allevieate this problem we have implemented the
techinuqe of linear interpolation of the rest shape mesh. At the start of
the simulation we use different rest shape mesh. It is topologicaly
equivalent to the original mesh but nodes have positions equivalent to the
simulated mesh (that way the initial deformations are not so big). During
the simulation we then perform several steps of linear interpolation
between these two rest shape meshes. At the end of interpolation the
original rest shape is used. After the mesh relaxes the simulation again
terminates.

% vim: et sw=2 tw=75 fdm=marker fdc=2 spell
